Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden

### 2023-05-12 analysis/paper/paper.qmd
Updated manuscript and restructured

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/archaeological-and-anthropological-sciences.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite has been a central material type utilized among hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period and is thus important to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of the same these materials. 125 points and preforms made from quartz/quartzite material were sampled from 46 archaeological sites located along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin and potentially the metamorphic grade. These features indicate the presence of graphite, muscovite, and biotite. Apart from reaffirming the potential of field-based screening instrumentation, these results should aid in future surveys of geological sources in the region. There also seems to be a good ground for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model could prove useful in spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
highlights: |
  These are the highlights. 
---

@@

Highlights: `r rmarkdown::metadata$highlights`

@@

Petrographic characterization and provenance studies are instrumental for inferences of trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region naturally raises important questions regarding the processes that resulted in its eventual deposition and discovery. The important first step in reconstructing the *chaîne opératoire*, as well as inferences of strategies for raw material selection and management, is the characterization of the material and identifying likely geological sources. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. The last decades, however, have seen an increase in the interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19]. The current bias that exists in the study of raw material management in regions where local flint and obsidian sources are non-existent has been highlighted by Prieto et al. [-@prieto_etal19: 15]. In Northern Sweden both quartz and quartzite dominate the lithic assemblages and have been used in studies of settlements, exchange and mobility [@baudou78]. While quartz and quartzite technology have been the target of several studies over the years, there has been less development in the petrographic characterization and provenience studies of these same materials [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. M While models of raw material sourcing and trade and exchange of slate and flint within Northern Sweden haves been put forth [@halen94: 115-122; @lundberg97: 161-171]. On the other hand there haves been little development on the topic of quartz and quartzite [@holm91: 23-24]. A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. The abundance of moraine deposits and the degree to which geological materials have been mixed and transported necessitates considerable effort to reconstruct the events that have formed the current landscape. Existing theories on lithic material sourcing within the North Swedish region tend to describe strategies regarding direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site from exposed veins in the bedrock, or as loose boulders and pebbles in the numerous moraine formations, as part of a strategy leaning more towards an embedded adaptation [@binford79; @rankama_etal06: 249-250]. Due to this ambiguous nature of availability of raw materials, provenance studies are complex.

Prieto et al. [-@prieto_etal19: 15; -@prieto_etal20: 32] notes that the usage of the term *quartzite* in archaeology has been somewhat ambiguous, often used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. It is difficult to provide a comprehensive definition of quartzite, owing much to its metamorphic nature. Because of this , and different researchers have thus emphasized different different features as being characteristics of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures, but Howard [-@howard05: 708] also emphasises the need for field-based methodologies. Spectroscopic surface measurements are fast, and can be adapted for use in the field. In the case of archaeological research there are situations where destructive techniques, such as thin-section analysis, are not suitable or applicable. Artefact collections (museum, private, etc.) represent a valuable resource and asset, with easy access to large datasets with a wide geographical distribution.

X-Ray Fluorescence (XRF) is a common spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) goes almost as far back as the XRF, as described by Norris [-@norris96], and there is an impressive body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

@@

In the current study a dataset of quartz and quartzite from Vilhelmina and Åsele parishes in Västerbotten, Sweden, have been analysed using the three above mentioned spectroscopic instrumentations (NIRS, Raman and XRF). The study includes points and preforms collected from `r length(tot_uniq_sites)` sites located along the Ångermanälven river valley, which runs through the western part of Västerbotten County (@fig-distribution-map). The aim is to assess the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region, as well as the material, necessitates alternative approaches to studies aimed at raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different "classes" of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and thus this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden during the 4 000 - 2 000 BP period.

@@

The current study includes spectroscopic surface measurements collected on quartz and quartzite points recovered from excavated and surveyed sites, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map). The sites are distributed along the Ångerman river valley, from the Lake Kultsjön in the north towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored at the collections at Västerbotten museum (www.vbm.se) have been analysed using NIRS, Raman scattering and XRF. No distinction is made between arrow- and spearheads.

A number of these points have, according to prior classifications, been worked using bifacial technology. The geographically closest chronology for bifacial points can be found in southeast Norway [@mjaerum12], but so far few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89], ). These attempts suggests a period of use between ca. 3 800 - 2 000 BP [@kaeck09: 60]. Similarly, studies of raw material use at sites dating to this period seems to indicate an increase in the use of quartzite when compared to prior periods [@baudou78].

@@

```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden.The glacial disturbance is visible in the widely distributd moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet."

@@

The dataset includes three main groups of registered petrographic material: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Those that had no prior classification available were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study serves two purposes. The first being to test the potential for separation of quartz from quartzite based on non-destructive analysis of their chemical and structural characteristics, and the second related purpose is to account for potential mistakes in the original classification.

When analysing the lithic material from the *Early Norrland* project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, however, the fragmented configuration of quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a somewhat common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91: 24]. As the material frequently has been characterized based on macroscopic characteristics, despite the central features necessary for characterization being microscopic, there is value in including a test group in the study.

@@

```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 2-76 mm. "

@@

All measurements were performed at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being the computer screens. The dataset consists of measurements collected on a heterogeneous material, not just in terms of colour, but also shape (@fig-artefacts - @fig-artefact-size) (see Supplementary Materials). The study is non-destructive sampling was performed on the surface of a material that is not uniform. The material ranges from fairly small and flat point fragments, to large and rough preforms. Therefore, the signal integration could at times be difficult, which in turn could lead to scattering effects and saturation.

Each artefact was sampled twice, on different sides, with additional measurements in cases where sampling was difficult. Sampling materials that featured high heterogeneity the sample point was focused on the surrounding matrix, away from deviating inclusions. An exception had to be made in the case of the XRF, however, where due to time constraints one measurement was collected as a rule apart from a few exceptions. This lowers the representativeness of the XRF analysis and inferences of the results should be evaluated with this in mind. The size of the dataset alleviates some of these issues, as for the larger sites there are multiple objects that may come from the same geological source, but the same can not be said for sites with smaller assemblages.

The artefacts features a mainly black or gray hue, as well as a subset of semi-translucent material with a beige tint, with some being almost fully translucent. While the colour of the material has been classified according to the Munsell system, for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent). In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.

@@

```{r}
#| label: fig-artefact-size
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.618
#| fig-align: center
#| fig-cap: "Scatter plot of length and width measurements, given in millimeters, for the sampled artefacts. 8 objects have no recorded thickness, which ranges between 3-76 mm. Note that the term *point* does not differentiate between spearheads or arrowheads."
quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
#knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)

@@

### Near Infrared {-}
NIRS measures the absorbance of light within the near infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near infrared region. The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify the bond. Overtones occur at frequencies that are a multiple of the infrared vibration. Most of them end up in the near infrared region [@osborne_etal93].

Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508].

NIR analysis was conducted with the Analytical Spectral Device (ASD) LabSpec 4, using a contact probe. The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before PCA modelling.

@@

### Raman {-}
Raman spectroscopy shares some similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared spectrum [@nafie01]. However, where NIRS measures light absorbance and generate overtones of molecular vibration, Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

### X-Ray Fluourescence {-}
XRF spectroscopy, and its capabilities for quantifying chemical elements in different materials, has made it a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Spectral features in XRF spectrometry can be related to certain elements according to Moseley's law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24]. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects that are caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.

XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer, connected to a portable test stand. The reference calibration used for element quantification was *mining* mode [@knight_etal21: 2]. The integration time was set to 130 seconds in order to cycle through all of the filters (main, low, high, light). Due to time constraints, and the number of artefacts included in the project, it was only possible to sample each object once, apart from some exceptions.

As the variables in XRF analysis is given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the *closed sum* effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods, and is managed by transforming the data using log-ratio transformation [@aitchison99]. There are currently three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. This makes multivariate methods like PCA suitable for identifying geochemical trends related to the rock-forming process.

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or "missing values") during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Since missing values are not permitted in the PCA these had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for "Log-ratio Expectation-Maximisation", which is an algorithm used to impute "left-censored data" (e.g. values below LOD, rounded zeros). Following log-transformation the data was then mean-centered and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].

@@

## PCA {-}

The data was evaluated using PCA, which is an unsupervised exploratory technique to reduce the dimensionality of a dataset [@geladi_linderholm20]. Multivariate approaches, such as PCA,  were proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predictive models.

@@

@@

## NIR - near infrared range {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

@@

@@

## XRF {-}

@@

@@

# Appendix A {#appendix-a}

Collection of loadings for each single instrumentation PCA.

\beginsupplement

>>>>

---
title: "Quartzite Complexities: Non-destructive analysis of bifacial points from Västerbotten, Sweden"
author:
  - Mattias Sjölander:
      email: mattias.sjolander@umu.se
      institute: [UmU]
      correspondence: true
  - Johan Linderholm:
      email: johan.linderholm@umu.se
      institute: [UmU]
      correspondence: false
  - Paul Geladi:
      email: paul.geladi@slu.se
      institute: [SLU]
      correspondence: false
  - Philip I. Buckland:
      email: philip.buckland@umu.se
      institute: [UmU]
      correspondence: false
institute:
  - UmU: Department of Historical, Philosophical and Religious studies, Umeå University, Umeå, Sweden
  - SLU: Department of Forest Biomaterials and Technology, Swedish University of Agricultural Sciences, Umeå, Sweden 
title-block-published: "Last updated"  
date: now
date-format: long
crossref: 
  fig-title: '**Fig.**'
  fig-labels: '**arabic**'
  fig-prefix: Fig.
format:
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 300
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/quaternary-science-reviews.csl" # Insert path for the bib-style
abstract: |
  Northern Fennoscandia represents a geologically complex region affected by both glacial and postglacial processes. Provenance of locally available geological material thus becomes complex and fraught with uncertainty. Quartzite was a key material type utilized by hunter-gatherers in Northern Sweden around the 4 000 – 2 000 BP period, and is thus critical to the understanding of raw material procurement and material flow within the region. However, there is a severe lack of methodological development in the characterization of these materials. 125 points and preforms made from quartz/quartzite were sampled from 46 archaeological sites along the upper Ångerman river valley in Västerbotten, Sweden. The material has been analysed non-destructively using three separate portable spectroscopic instrumentations (Near-infrared, Raman, X-Ray Fluorescence). Evaluation of the spectra and exploratory data analysis using Principal Component Analysis demonstrates detectable differences in the material that likely stem from diagenetic/paragenetic origin. The presence of graphite, muscovite and biotite could likewise provide information on the material’s metamorphic grade. In addition to reaffirming the potential of field-based screening instrumentation, these results will benefit future surveys of geological sources in the region. They also indicate potential for the construction of a predictive model that could classify the quartzite based on its chemical characteristic. Such a model would prove useful in future spatial analysis and testing of models of raw material management.
keywords: |
  Lithics; Provenance; Quartzite; Spectroscopy; Sweden
---

@@

Highlights:

* Non-destructive analysis of prehistoric bifacial points made in quartzite using near-infrared spectroscopy, Raman scattering and X-Ray Fluourescence. 
* Each method detects differences in the structural and chemical characteristics in the material, but on their own the resolution is too low for predictive modelling. 
* Data fusion techniques are recommended for predictive modelling using non-destructive spectroscopy on quartzite material.

@@

Petrographic characterization and provenance studies are instrumental for inferring trade and exchange among prehistoric societies [@tykot03]. The presence of lithic material foreign to the local region of an archaeological site, raises important questions on the processes that resulted in its eventual deposition and discovery. The characterization of geological material and identification of geological sources are the first steps in reconstructing a prehistoric community’s strategies for raw material selection and management. Flint and obsidian have been central to studies of trade and exchange in European prehistory, due in part to the well-surveyed geological sources for these materials [@tykot17]. This focus has led to a bias, however, which has been highlighted by Prieto et al. [-@prieto_etal19: 15]. Recent decades, however, have seen an increase in interest for non-flint lithics and raw material studies [@blomme_etal12; @dalpra_pitblado16; @prieto_etal19; @prieto_etal20; @prieto_etal21; @ramacciotti_etal19].

In Northern Sweden, quartz and quartzite dominate the lithic assemblages and have been used in settlement, exchange and mobility studies [@baudou78]. Models have been proposed for raw material sourcing and trade and exchange for the less common slate and flint finds within Northern Sweden [@halen94: 115-122; @lundberg97: 161-171]. There has, however, been little development in the petrographic characterization and provenance study of quartz and quartzite [@broadbent79; @callahan_etal92; @knutsson_etal16; @tallavaara_etal10]. ). A possible reason for this might well be the complex geological background of the region [@bargel03; @stephens_bergmanweihed20; @stroeven_etal16]. Northern Sweden is extensively covered by glacial till, and whilst there are identifiable source areas (e.g. small bedrock quarries), there is an extremely large number of other potential sources of raw materials to be found in moraine deposits. The few existing theories on lithic material sourcing within the North Swedish region describe strategies for direct procurement from the mountain zone [@forsberg85; -@forsberg12]. This does not, however, account for the possibility that the material was sourced locally around the site, as part of an embedded adaptation strategy [@binford79; @rankama_etal06: 249-250]. A new approach to provenance studies is therefore required in this region.

Prieto et al. [-@prieto_etal19: 15; -@prieto_etal20: 32] note the ambiguous usage of the term quartzite in archaeology, with it often being used interchangeably between sedimentary orthoquartzite and metamorphic quartzite. When defining quartzite, different researchers have emphasised different features as being characteristics of the material [@howard05]. Visual characterization with the naked eye is typically unreliable, and microscopic analysis is recommended for the identification of metamorphic structures. The need for field-based methodologies is, however, stressed by Howard [-@howard05: 708]. Spectroscopic surface measurements are fast and can be adapted for use in the field. However, artefact collections (museum, private, etc.) also represent a valuable resource and asset, with easy access to large potential datasets with a wide geographical distribution. Destructive analysis on material from these collections, which are often unique and include sensitive cultural heritage material, are not always suitable or permitted, and thus there is a clear need for the development and testing of non-destructive techniques such as those presented in this study.

In this study, three different spectroscopic techniques have been applied to material stored in collections at Västerbottens museum in Umeå, Sweden. X-Ray Fluorescence (XRF) is a well-established spectroscopic technique for petrographic characterization [@shackley11a; @tykot03], due to its capability of providing element compositional data. The use of near-infrared spectroscopy (NIRS) is similarly well-known, as described by Norris [-@norris96], and there is a large body of work dedicated to the fingerprinting of geological materials [@hunt77; @clark_etal90; @linderholm_geladi14a; @linderholm_geladi14b; @sciuto18]. Complementary to NIR, Raman spectroscopy has been applied extensively in both the Cultural Heritage sector, in the analysis of artwork and manuscripts, as well as the Earth sciences [@dubessy_etal12].

@@

This paper presents an assessment of the potential for non-destructive characterization of quartz and quartzite material without the use of references from known geological sources. The geological complexity of both the region and the material necessitates alternative approaches to studies of raw material management. Without relying on well-delimited source areas for geological material, a predictive model capable of separating different “classes” of quartzite would enable spatial analysis of settlement and mobility, as well as testing of established models. The currently available data on the chemical characteristic of quartz and quartzite in this region of Northern Sweden is limited, however, and this study represents a first step towards developing the topic of raw material management of these materials in Northern Sweden between 4 000 and 2 000 BP.

@@

This study includes spectroscopic surface measurements collected on bifacial points (probable arrowheads and spear points) made of quartz and quartzite. The points have been recovered from archaeological sites during excavations and surveys, including stray finds, in Västerbotten County, Sweden (@fig-distribution-map). The sites are distributed along the Ångerman river valley, from Lake Kultsjön in the north, towards the town of Åsele in the south. The dataset consists of `r length(tot_uniq_samples)` points, point fragments and preforms stored in the collections at Västerbottens museum (www.vbm.se) and have been analysed using NIRS, Raman scattering and XRF.

The geographically closest chronology for bifacial points is centred around southeast Norway [@mjaerum12], but a few attempts have been made in the North Swedish region [@baudou92: 99; @forsberg85: 5; @forsberg89]. The latter suggest a period of use between ca. 3 800 - 2 000 BP [@kaeck09: 60]. Studies of raw material use at sites dating to this period seem to indicate an increase in the use of quartzite when compared to earlier periods [@baudou78].

@@

```{r}
#| label: fig-distribution-map
#| cache: false
#| cache-lazy: false
#| fig-cap: "Geographical distribution of the sampled sites along the river of Ångermanälven, Västerbotten, Sweden. The glacial disturbance is visible in the widely distributed moraine. Tectonic symbology after [@gee_stephens20]. Made with © Sveriges Geologiska Undersökning, Lantmäteriet (CC0), Natural Earth, and Riksantikvarieämbetet"

@@

The dataset includes three main groups of petrographic material, as classified by archaeologists: quartz, quartzite, and brecciated quartz (Swe. "*brecciekvarts*"). Artefacts that had no prior classification were sorted into either the quartz or quartzite category based on visual inspection. This resulted in: `r length(bquartz)` brecciated quartz samples, `r length(quartz)` quartz samples, and `r length(quartzite)` quartzite samples. The inclusion of quartz and brecciated quartz in the study enables the separation of quartz from quartzite, based on non-destructive analysis of their chemical and structural characteristics, to be evaluated.

In his analysis of the petrographic material in the Early Norrland project [@biornstad68; @ahman67], the geologist Åhman [-@ahman67: 8] introduced the term *brecciated quartz* (also referred to as *mylonite quartz*) to archaeologists in Northern Sweden. Åhman describes the material as easily mistaken for an unusually pure quartzite, but where the fragmented configuration of the quartz can be identified via microscopic analysis. *Brecciated quartz* in particular became a common term used in North Swedish archaeology, and has typically been used to identify a certain translucent type of fine-grained, quartz-rich material [e.g. @holm91: 24]. As the material frequently has been characterized on the basis of macroscopic characteristics, despite the fundamental features necessary for characterization being microscopic, there is value in including this material as a test group in the study.

@@

```{r}
#| label: fig-artefacts
#| cache: false
#| cache-lazy: false
#| fig-cap: "Examples of the bifacial points studied. The material is heterogeneous in terms of colour, and the surface of the artefacts themselves is not uniform. Ranging from near perfect points to rough preforms, it can be difficult to sample due to light scattering effects and saturation. Length varies between 12 – 138 mm, width between 8 – 67 mm, and thickness between 3 – 76 mm"

@@

All measurements were collected at the Environmental Archaeology Lab at Umeå University (www.umu.se/en/research/infrastructure/mal/). The analysis was conducted in a dark room in order to minimize stray light, with the main light sources in the room being 2 computer screens. The sampled material is heterogeneous in colour and shape (@fig-artefacts - @fig-artefact-size) (see supplementary data). The artefacts range from small and flat point fragments, to large and rough preforms. Therefore, the signal integration could at times be difficult, leading to scattering effects and saturation. For NIR and Raman analyses, each artefact was sampled twice, on different sides, with additional measurements where sampling was difficult. When sampling the probe was in contact with the surface of the artefact at as flat a surface as possible. If the material featured high heterogeneity, the sample point was focused on the surrounding matrix, avoiding deviating inclusions (e.g. the dark mineral veins in @fig-artefact-size).

For XRF, the more time-consuming nature of the measurement process necessitated only one measurement per object. The representativeness of the XRF analysis is thus lower, and inferences derived from the results in this study should be used with caution. The risk of misinterpretation is, however, somewhat compensated by the large size of the dataset, with larger sites including multiple objects that may come from the same geological source. The results from sites with smaller assemblages are, however, more liable to be affected by the limited range of material analysed.

In colour, the artefacts exhibit a mainly black or grey hue, with a subset of semi-translucent material with a beige tint, with some being almost fully translucent. While the colour of the material has been classified according to the Munsell system (see supplementary data), for visualization purposes they have also been sorted into general hues: *dark* (~ N1-N4), *light* (~ N5-N8), *white* (~ N9), and *colourless* (lack of colour, opaque to translucent). In the case of heterogeneous objects the Munsell colour and general hue describes the surrounding matrix.

@@

```{r}
#| label: fig-artefact-size
#| cache: true
#| fig-width: 7.87
#| fig-asp: 0.618
#| fig-align: center
#| fig-cap: "Scatter plot of length and width measurements (mm) for the sampled artefacts. 8 artefacts have no recorded thickness, which ranges between 3-76 mm. Note that the term point does not differentiate between spear points or arrowheads"
quietly(source(here::here("analysis", "paper", "001-artefact-size.R"), encoding = "UTF-8"))
#knitr::include_graphics(here::here("analysis", "figures", "001-artefact-size.jpeg"), error = FALSE)

@@

### Near Infrared Spectroscopy (NIRS) {-}

Many chemical bonds between small atoms (H, C, O, N etc.) show a fundamental vibration in the near-infrared region (780 nm – 2 500 nm) of the electromagnetic spectrum. The vibration occurs at a frequency that is typical for the bond and can be used to identify and quantify it. Based on the theory of molecular vibration and excitation [@hunt77], by exposing a geological material to electromagnetic radiation it is possible to generate overtone vibrations which can be related to certain molecular groups. The most common group capable of generating these overtones is the OH group, which in quartz-based materials typically relates to molecular water and metal-OH combinations [@hunt77: 508]. Overtones occur at frequencies that are a multiple of the infrared vibration, and most of these end up in the near-infrared region [@osborne_etal93].

NIR analysis was conducted with an Analytical Spectral Device (ASD) LabSpec 4, using a contact probe (spot size: 10 mm). The ASD features a detection range of 350 nm – 2 500 nm, with spectral sampling (resp. spectral resolution) of 1.4 nm (resp. 3 nm) in the visible and near infrared range and 1.1 nm (resp. 10 nm) in the short‐wave infrared range. Each object was measured against a white reference background and the collected measurements were averaged for each artefact. The data was restricted to the 1 000 nm - 2 500 nm range and mean-centred before Principal Component Analysis (PCA) modelling.

@@

### Raman {-}
Raman spectroscopy shares similarities with NIRS in that they both can be used to infer molecular structure based on their vibrational properties within the near-infrared region [@nafie01]. Raman spectroscopy is based on the *Raman scattering* effect. By exposing a molecule to radiation with a monochromatic laser it is possible to induce a dipole moment in the molecule which generates the Raman effect. The subsequent light that is scattered contains both Rayleigh and Raman scatter, the former at a frequency of incident radiation and the latter at a shifted frequency [@dubessy_etal12]. The spectrometer will measure the Raman scatter and generate a spectrum where the spectral line has shifted compared to the incident Rayleigh frequency. As the structure of a molecule will inform its vibrational mode, the point at which these shifts occur within the infrared region varies depending on the material [@smith_carabatos-nedelec01].

@@

### X-Ray Fluorescence (XRF) {-}
XRF spectroscopy is a well-established method in petrographic characterization in archaeology [@hall60; @shackley11a]. Spectral features in XRF spectrometry can be related to certain elements according to Moseley’s law [@moseley1913], which recognizes that the energy of the fluorescence produced when a material is irradiated with high energy particles is characteristic to certain atoms [@thyrel14: 20-24]. Quantification based on these features is complex, however, and a number of methods have been developed as a result [@sitko_zawisa12: 143-159]. In a non-destructive context, where the sample is not homogeneous, it is necessary to correct for matrix effects caused by the presence of other elements in a sample. One way to do so is calibration using pure-element standards, with which a calibration constant can be calculated.
XRF measurements were collected using a Thermo Scientific Niton XL5 Analyzer (spot size: 8 mm), connected to a portable test stand. The reference calibration used for element quantification was mining mode [@knight_etal21: 2]. The integration time was set to 130 seconds.

As the variables in XRF analysis are given as proportions of a constant sum (e.g. percentage, parts per million (ppm)), this creates something referred to as the closed sum effect [@pawlowsky-glahn_egozcue06]. This means that if one element was to change in value, others would follow to maintain the constant [@grunsky_caritat20]. This has implications for classical statistical methods and is managed by transforming the data using log-ratio transformation [@aitchison99]. Currently there are three main methods; additive log-ratio (alr), centered log-ratio (clr) and isometric log-ratio [@pawlowsky-glahn_etal15]. In the PCA, the Principal Components are orthogonal and thus the relationships between variables treated with log-ratio transformation are linear and independent of the total concentration [e.g. @reimann_etal12]. With this transformation, PCA is thus suitable for exploring geochemical trends related to the rock-forming process without being impacted by the closed sum effect (REF).

As the current dataset features values below the Limit of Detection (LOD) these were treated as NULL (or “missing values”) during PCA. A LOD value is defined as a measured value 3 times lower than the standard deviation for that measurement. Missing values are not permitted in the PCA and thus had to be imputed before the analysis, which was performed using the *lrEM* function from the *zCompositions* R package [@palarea-albaladejo_martin-fernandez15]. *lrEM* stands for “Log-ratio Expectation-Maximisation”, which is an algorithm used to impute “left-censored data” (e.g. values below LOD, rounded zeros). Following log-transformation the data was mean-centred and scaled to unit variance before PCA. Samples that were less than 10 mm in length or width were excluded from the model [@shackley11b: 9].

@@

## Principal Component Analysis (PCA) {-}

The data were evaluated using PCA, an unsupervised exploratory technique for reducing the dimensionality of a dataset  [@geladi_linderholm20]. Multivariate approaches, such as PCA, have been proven useful for exploring complex spectroscopic datasets [@sciuto18]. Clusters and trends that emerge in the resulting score plot can potentially be related to the spectral fingerprints of the material. These groupings can then be followed up with supervised methods, and aid in the construction of predictive models for classifying unknown materials according to previously defined groups.

@@

Spectra and PCA loadings for each instrumentation can be found in the supplementary data. An open repository with the code used in the analysis is linked in the *Data availability* section.

@@

## Near Infrared Spectroscopy (NIRS) {-}
```{r nir-sample-count, results='hide'}
#Select all dark samples
points.d.tot <-
  metadata.csv %>%
  filter( type == "Point" | type == "Point fragment" | type == "Preform", 
         material == "Brecciated quartz" | material == "Quartz" | material == "Quartzite") %>% 
  filter(hue == "Dark")

@@

fig
```
-->

@@

fig
```
-->
 \linebreak

@@

## X-Ray Fluorescence (XRF) {-}

@@

Philip I. Buckland: Writing – original draft, Writing - Review & Editing.

@@
